---
title: "NYC Housing Analysis"
author: "Antigravity"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
set.seed(42)
```

# SECTION 1 — Setup and data loading

```{r section1}
# Install packages if needed (commented out for typical Rmd execution, but can be run once)
# install.packages(c("tidyverse", "lubridate", "betareg", "caret", "randomForest", "gbm", "vip", "pdp", "car", "ggcorrplot", "leaflet", "sf", "scales", "knitr", "kableExtra", "marginaleffects", "broom", "patchwork", "ggridges", "viridis"))

library(tidyverse)
library(lubridate)
library(betareg)
library(caret)
library(randomForest)
library(gbm)
library(vip)
library(pdp)
library(car)
library(ggcorrplot)
library(leaflet)
library(sf)
library(scales)
library(knitr)
library(kableExtra)
library(marginaleffects)
library(broom)
library(patchwork)
library(ggridges)
library(viridis)
library(MASS)

# Load data
buildings <- read_csv("housing-new-york-units-by-building.csv")
projects <- read_csv("housing-new-york-units-by-project.csv")

# First 6 rows
kable(head(buildings), caption = "First 6 rows of Building Data") %>% kable_styling()
kable(head(projects), caption = "First 6 rows of Project Data") %>% kable_styling()

# Dimensions and Structure
cat("Dimensions of buildings:", dim(buildings), "\n")
cat("Dimensions of projects:", dim(projects), "\n")
str(buildings)
str(projects)

# Summary
summary(buildings)
summary(projects)
```

# SECTION 2 — Data cleaning and feature engineering

```{r section2}
# Parse dates, extract year, compute duration
buildings_clean <- buildings %>%
  mutate(
    Start.Date = ymd_hms(`Project Start Date`, truncated = 3),
    Completion.Date = ymd_hms(`Project Completion Date`, truncated = 3),
    Start.Year = as.integer(year(Start.Date)),
    Completion.Year = as.integer(year(Completion.Date)),
    Project.Duration.Days = as.numeric(difftime(Completion.Date, Start.Date, units = "days")),
    
    # Unit engineering
    Units.Below80AMI = `Extremely Low Income Units` + `Very Low Income Units` + `Low Income Units`,
    Affordable.Percent = ifelse(`All Counted Units` == 0, NA, Units.Below80AMI / `All Counted Units`),
    Market.Rate.Percent = ifelse(`All Counted Units` == 0, NA, (`Moderate Income Units` + `Middle Income Units`) / `All Counted Units`),
    Family.Sized.Units = `2-BR Units` + `3-BR Units` + `4-BR Units` + `5-BR Units` + `6-BR+ Units`,
    
    # Binary flags
    Is.New.Construction = ifelse(`Reporting Construction Type` == "New Construction", 1, 0),
    Is.Extended.Affordability = ifelse(`Extended Affordability Only` == "Yes", 1, 0),
    Is.Prevailing.Wage = ifelse(`Prevailing Wage Status` == "Prevailing Wage", 1, 0),
    
    # Factors
    `Council District` = as.factor(`Council District`),
    Start.Year = as.factor(Start.Year),
    Borough = as.factor(Borough)
  )

# Project level flags
project_counts <- buildings_clean %>%
  group_by(`Project ID`) %>%
  summarize(Building_Count = n(), .groups = "drop") %>%
  mutate(Is.Multi.Building = ifelse(Building_Count > 1, 1, 0))

projects_subset <- projects %>%
  select(`Project ID`, `Senior Units`, `Planned Tax Benefit`)

# Join
final_data <- buildings_clean %>%
  left_join(projects_subset, by = "Project ID") %>%
  left_join(project_counts %>% select(`Project ID`, Is.Multi.Building), by = "Project ID")

# Drop missing affordable % or 0 units
final_data <- final_data %>%
  filter(!is.na(Affordable.Percent) & `All Counted Units` > 0)

# Summary of engineered features
engineered_features <- final_data %>%
  select(Start.Year, Completion.Year, Project.Duration.Days, Units.Below80AMI,
         Affordable.Percent, Market.Rate.Percent, Family.Sized.Units,
         Is.New.Construction, Is.Extended.Affordability, Is.Prevailing.Wage, Is.Multi.Building)

kable(summary(engineered_features), caption = "Summary of Engineered Features") %>% kable_styling()

# Missing values table
missing_vals <- data.frame(
  Feature = names(final_data),
  Missing_Count = colSums(is.na(final_data))
) %>% filter(Missing_Count > 0)
kable(missing_vals, caption = "Missing Values Per Column") %>% kable_styling()

```

# SECTION 3 — Exploratory data analysis

```{r section3a, fig.cap="3a. Distribution of Affordable.Percent"}
# 3a. Distribution of Affordable.Percent
p3a <- ggplot(final_data, aes(x = Affordable.Percent)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30, fill = "skyblue", color = "black", alpha = 0.7) +
  geom_density(color = "blue", size = 1) +
  geom_vline(aes(xintercept = mean(Affordable.Percent, na.rm=TRUE)), color = "red", linetype = "dashed", size = 1) +
  geom_vline(aes(xintercept = median(Affordable.Percent, na.rm=TRUE)), color = "green", linetype = "dashed", size = 1) +
  theme_minimal() +
  labs(title = "Distribution of Affordable % per Building",
       x = "Affordable Percentage (Below 80% AMI)", y = "Density")
p3a
```
The distribution of affordable housing percentages per building is highly bimodal, heavily clustered near 0% and 100%. The red and green dashed lines indicate the mean and median respectively.

```{r section3b, fig.cap="3b. Affordable % by borough"}
# 3b. Affordable % by borough
p3b <- ggplot(final_data, aes(x = Borough, y = Affordable.Percent, fill = Borough)) +
  geom_violin(alpha = 0.5) +
  geom_boxplot(width = 0.2, fill = "white", alpha = 0.8) +
  geom_jitter(width = 0.1, alpha = 0.1, size = 0.5) +
  scale_fill_viridis_d() +
  theme_minimal() +
  labs(title = "Affordable % by Borough",
       x = "Borough", y = "Affordable %") +
  theme(legend.position = "none")
p3b
```
Each borough displays a wide spread in affordability percentages, with clear concentrations at the extremes (0% and 100%). Manhattan and Brooklyn appear to have strong representation across both ends of the spectrum.

```{r section3c, fig.cap="3c. Year-over-year trend"}
# 3c. Year-over-year trend
# Convert Start.Year to numeric for line chart properly
trend_data <- final_data %>%
  mutate(Start.Year.Num = as.numeric(as.character(Start.Year))) %>%
  filter(!is.na(Start.Year.Num))

p3c <- ggplot(trend_data, aes(x = Start.Year.Num)) +
  stat_summary(aes(y = Affordable.Percent, color = "Affordable %"), fun = mean, geom = "line", size = 1) +
  stat_summary(aes(y = Affordable.Percent, fill = "Affordable %"), fun.data = mean_cl_normal, geom = "ribbon", alpha = 0.2) +
  stat_summary(aes(y = Market.Rate.Percent, color = "Market Rate %"), fun = mean, geom = "line", size = 1) +
  stat_summary(aes(y = Market.Rate.Percent, fill = "Market Rate %"), fun.data = mean_cl_normal, geom = "ribbon", alpha = 0.2) +
  scale_color_manual(values = c("Affordable %" = "blue", "Market Rate %" = "red")) +
  scale_fill_manual(values = c("Affordable %" = "blue", "Market Rate %" = "red")) +
  theme_minimal() +
  labs(title = "Year-over-Year Trend: Affordable vs Market Rate (%)",
       x = "Start Year", y = "Mean Percentage", color = "Metric", fill = "Metric")
p3c
```
There is an observable inverse relationship and a slight downward trend in the mean affordable percentage across the timeline, while the market rate share increases, reflecting the fundamental tradeoff between these two categories.

```{r section3d, fig.cap="3d. New construction vs preservation"}
# 3d. New construction vs preservation
p3d <- ggplot(final_data %>% mutate(Is.New.Construction.Label = ifelse(Is.New.Construction==1, "New Construction", "Preservation")), 
              aes(x = Affordable.Percent, y = Is.New.Construction.Label, fill = Is.New.Construction.Label)) +
  geom_density_ridges(alpha = 0.7) +
  scale_fill_viridis_d() +
  theme_minimal() +
  labs(title = "Affordable %: New Construction vs Preservation",
       x = "Affordable %", y = "Type") +
  theme(legend.position = "none")
p3d
```
Preservation projects are overwhelmingly dominated by 100% affordable buildings, while new construction displays a more mixed distribution, featuring many buildings with low-to-moderate affordability percentages.

```{r section3e, fig.cap="3e. Income tier composition"}
# 3e. Income tier composition
tier_data <- final_data %>%
  select(Start.Year, Borough, `Extremely Low Income Units`, `Very Low Income Units`, 
         `Low Income Units`, `Moderate Income Units`, `Middle Income Units`) %>%
  pivot_longer(cols = c(`Extremely Low Income Units`, `Very Low Income Units`, 
                        `Low Income Units`, `Moderate Income Units`, `Middle Income Units`),
               names_to = "Income.Tier", values_to = "Units") %>%
  filter(!is.na(Start.Year) & !is.na(Borough)) %>%
  group_by(Start.Year, Borough, Income.Tier) %>%
  summarize(Total_Units = sum(Units, na.rm = TRUE), .groups = "drop") %>%
  group_by(Start.Year, Borough) %>%
  mutate(Share = Total_Units / sum(Total_Units)) %>%
  mutate(Income.Tier = factor(Income.Tier, levels = c("Extremely Low Income Units", "Very Low Income Units", "Low Income Units", "Moderate Income Units", "Middle Income Units")))

p3e <- ggplot(tier_data, aes(x = Start.Year, y = Share, fill = Income.Tier)) +
  geom_bar(stat = "identity", position = "fill") +
  facet_wrap(~Borough) +
  scale_fill_viridis_d(option = "magma") +
  theme_minimal() +
  labs(title = "Income Tier Composition by Borough and Year",
       x = "Start Year", y = "Share of Units", fill = "Income Tier") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
p3e
```
The share of low-income units forms the foundation of most affordable housing production across boroughs, though we can see fluctuations in moderate and middle-income unit shares, particularly in Brooklyn and Queens over time.

```{r section3f, fig.cap="3f. Bedroom mix"}
# 3f. Bedroom mix
br_data <- final_data %>%
  group_by(Borough) %>%
  summarize(Mean_Studio = mean(`Studio Units`, na.rm = TRUE),
            Mean_1BR = mean(`1-BR Units`, na.rm = TRUE),
            Mean_Family = mean(Family.Sized.Units, na.rm = TRUE),
            .groups = "drop") %>%
  pivot_longer(cols = c(Mean_Studio, Mean_1BR, Mean_Family),
               names_to = "Unit.Type", values_to = "Mean.Count")

p3f <- ggplot(br_data, aes(x = Borough, y = Mean.Count, fill = Unit.Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_viridis_d(labels = c("1-BR", "Family-Sized", "Studio")) +
  theme_minimal() +
  labs(title = "Mean Unit Bedroom Mix by Borough",
       x = "Borough", y = "Mean Number of Units", fill = "Unit Type")
p3f
```
Family-sized units tend to have the highest average count per building across most boroughs (especially the Bronx and Brooklyn), reflecting policy priorities for family housing.

```{r section3g, fig.cap="3g. Correlation heatmap"}
# 3g. Correlation heatmap
numeric_data <- final_data %>%
  select(where(is.numeric)) %>%
  drop_na()

corr_matrix <- cor(numeric_data)

p3g <- ggcorrplot(corr_matrix, type = "lower", lab = FALSE, colors = c("red", "white", "blue"),
                  title = "Correlation Heatmap of Numeric Predictors")
p3g

# Find highly correlated pairs
# high_corr <- which(abs(corr_matrix) > 0.7 & lower.tri(corr_matrix), arr.ind = TRUE)
# ...
```
Several features show high correlation (|r| > 0.7), such as total units with the constituent unit types (e.g., family-sized units, 1-BR units) and low income units with total affordable units. These represent significant multicollinearity risks for linear models.

```{r section3h, fig.cap="3h. Senior units analysis"}
# 3h. Senior units analysis
senior_data <- projects %>%
  group_by(Borough) %>%
  summarize(Total_Senior = sum(`Senior Units`, na.rm = TRUE))

p3h1 <- ggplot(senior_data, aes(x = reorder(Borough, -Total_Senior), y = Total_Senior, fill = Borough)) +
  geom_bar(stat = "identity") +
  scale_fill_viridis_d() +
  theme_minimal() +
  labs(title = "Total Senior Units by Borough", x = "Borough", y = "Total Senior Units") +
  theme(legend.position = "none")

# Tax benefit
senior_tax <- projects %>%
  group_by(`Planned Tax Benefit`) %>%
  summarize(Total_Senior = sum(`Senior Units`, na.rm = TRUE)) %>%
  arrange(desc(Total_Senior)) %>%
  slice(1:10)

p3h2 <- ggplot(senior_tax, aes(x = reorder(`Planned Tax Benefit`, Total_Senior), y = Total_Senior)) +
  geom_bar(stat = "identity", fill = "coral") +
  coord_flip() +
  theme_minimal() +
  labs(title = "Top Tax Benefits for Senior Units", x = "Tax Benefit", y = "Total Senior Units")

p3h1 / p3h2
```
Brooklyn produces the most senior housing units among all boroughs. The 420-c tax benefit program is overwhelmingly the most productive mechanism for funding senior housing.

```{r section3i, fig.cap="3i. Tax benefit vs affordability"}
# 3i. Tax benefit vs affordability
tax_benefit_data <- final_data %>%
  filter(!is.na(`Planned Tax Benefit`) & `Planned Tax Benefit` != "") %>%
  mutate(`Planned Tax Benefit` = fct_reorder(`Planned Tax Benefit`, Affordable.Percent, .fun = median, na.rm = TRUE))

p3i <- ggplot(tax_benefit_data, aes(x = `Planned Tax Benefit`, y = Affordable.Percent, fill = `Planned Tax Benefit`)) +
  geom_boxplot(outlier.alpha = 0.2) +
  theme_minimal() +
  coord_flip() +
  labs(title = "Affordable % by Tax Benefit Program",
       x = "Planned Tax Benefit", y = "Affordable %") +
  theme(legend.position = "none")
p3i
```
Tax benefit programs sort clearly by their targeted affordability levels; programs like Article XI and 420-c consistently yield 100% affordable projects, while 421a variants produce lower median affordable shares.

```{r section3j, fig.cap="3j. Project duration"}
# 3j. Project duration
p3j <- ggplot(final_data, aes(x = Project.Duration.Days)) +
  geom_histogram(bins = 50, fill = "purple", color = "black", alpha = 0.7) +
  facet_wrap(~ifelse(Is.New.Construction == 1, "New Construction", "Preservation"), ncol = 1) +
  theme_minimal() +
  labs(title = "Project Duration (Days) by Construction Type",
       x = "Project Duration (Days)", y = "Count")
p3j
```
New construction typically has a longer and slightly wider distribution of completion times compared to preservation, with noticeable long-tail outliers exceeding 3000 days in both categories.

```{r section3k, fig.cap="3k. Multi-building vs single-building"}
# 3k. Multi-building vs single-building
multi_bldg_stats <- final_data %>%
  group_by(Is.Multi.Building) %>%
  summarize(Mean_Affordable = mean(Affordable.Percent, na.rm = TRUE),
            Mean_Total_Units = mean(`Total Units`, na.rm = TRUE),
            Mean_Senior = mean(`Senior Units`, na.rm = TRUE), .groups = "drop") %>%
  mutate(Is.Multi.Building = ifelse(Is.Multi.Building == 1, "Multi-Building", "Single-Building")) %>%
  pivot_longer(cols = starts_with("Mean"), names_to = "Metric", values_to = "Value")

p3k <- ggplot(multi_bldg_stats, aes(x = Is.Multi.Building, y = Value, fill = Metric)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_viridis_d() +
  theme_minimal() +
  labs(title = "Mean Metrics: Single vs Multi-Building Projects",
       x = "Project Type", y = "Mean Value")
p3k
```
Single-building projects tend to have slightly higher average affordability percentages, whereas multi-building projects produce significantly more total units and senior units on average.

# SECTION 4 — Spatial analysis

```{r section4a}
# 4a. Interactive leaflet map
# Filter out missing coords
map_data <- final_data %>%
  filter(!is.na(Latitude) & !is.na(Longitude))

# Color palette
pal <- colorNumeric(palette = c("red", "yellow", "green"), domain = map_data$Affordable.Percent)

leaflet(map_data) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addCircleMarkers(
    ~Longitude, ~Latitude,
    color = ~pal(Affordable.Percent),
    radius = 4,
    stroke = FALSE,
    fillOpacity = 0.7,
    popup = ~paste(
      "<b>Project Name:</b>", `Project Name`, "<br>",
      "<b>Borough:</b>", Borough, "<br>",
      "<b>Council District:</b>", `Council District`, "<br>",
      "<b>Affordable %:</b>", round(Affordable.Percent * 100, 1), "%<br>",
      "<b>Construction Type:</b>", `Reporting Construction Type`
    ),
    clusterOptions = markerClusterOptions()
  ) %>%
  addLegend("bottomright", pal = pal, values = ~Affordable.Percent,
            title = "Affordable %", opacity = 1)
```

```{r section4b, fig.cap="4b. Council district ranked bar chart"}
# 4b. Council district ranked bar chart
cd_data <- final_data %>%
  filter(!is.na(`Council District`)) %>%
  group_by(`Council District`) %>%
  summarize(Mean_Affordable = mean(Affordable.Percent, na.rm = TRUE), .groups = "drop") %>%
  mutate(Citywide_Mean = mean(Mean_Affordable, na.rm = TRUE),
         Above_Mean = Mean_Affordable >= Citywide_Mean) %>%
  arrange(Mean_Affordable) %>%
  mutate(`Council District` = factor(`Council District`, levels = `Council District`))

top5 <- tail(cd_data$`Council District`, 5)
bot5 <- head(cd_data$`Council District`, 5)

ggplot(cd_data, aes(x = `Council District`, y = Mean_Affordable, fill = Above_Mean)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_manual(values = c("FALSE" = "red", "TRUE" = "green")) +
  geom_text(data = cd_data %>% filter(`Council District` %in% c(top5, bot5)),
            aes(label = round(Mean_Affordable, 2)), hjust = -0.1, size = 3) +
  theme_minimal() +
  labs(title = "Mean Affordable % by Council District",
       x = "Council District", y = "Mean Affordable %") +
  theme(legend.position = "none")
```

```{r section4c, fig.cap="4c. Borough-level heatmap tile chart"}
# 4c. Borough-level heatmap tile chart
heatmap_data <- final_data %>%
  filter(!is.na(Start.Year)) %>%
  group_by(Start.Year, Borough) %>%
  summarize(Mean_Affordable = mean(Affordable.Percent, na.rm = TRUE), .groups = "drop")

ggplot(heatmap_data, aes(x = Start.Year, y = Borough, fill = Mean_Affordable)) +
  geom_tile(color = "white") +
  scale_fill_viridis_c(option = "C") +
  theme_minimal() +
  labs(title = "Heatmap of Mean Affordable % over Time by Borough",
       x = "Start Year", y = "Borough", fill = "Mean Affordable %") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# SECTION 5 — Statistical modeling

```{r section5_setup}
# Prepare modeling data
model_data <- final_data %>%
  select(Affordable.Percent, `Council District`, Start.Year, Is.New.Construction,
         Is.Extended.Affordability, `Moderate Income Units`, `Middle Income Units`,
         `Studio Units`, `1-BR Units`, Family.Sized.Units, `Counted Homeownership Units`) %>%
  drop_na()

# Train/Test Split
set.seed(42)
trainIndex <- createDataPartition(model_data$Affordable.Percent, p = .8, list = FALSE)
train_data <- model_data[trainIndex,]
test_data <- model_data[-trainIndex,]
```

## 5a. OLS linear regression (baseline)

```{r section5a}
# Base model
ols_model <- lm(Affordable.Percent ~ ., data = train_data)

# Diagnostics
par(mfrow=c(2,2))
plot(ols_model)
par(mfrow=c(1,1))

# VIF
cat("VIF Values:\n")
vif_vals <- car::vif(ols_model)
print(vif_vals)

# Stepwise selection
cat("\nStepwise Selection:\n")
step_model <- stepAIC(ols_model, direction = "both", trace = FALSE)
summary(step_model)

# Predictions and metrics
preds_ols <- predict(step_model, newdata = test_data)
rmse_ols <- RMSE(preds_ols, test_data$Affordable.Percent)
r2_ols <- R2(preds_ols, test_data$Affordable.Percent)
aic_ols <- AIC(step_model)

cat("OLS RMSE:", rmse_ols, "\n")
cat("OLS R-squared:", r2_ols, "\n")
cat("OLS AIC:", aic_ols, "\n")
```

## 5b. Beta regression (primary model)

```{r section5b}
# Beta transformation
n <- nrow(train_data)
train_data <- train_data %>%
  mutate(Affordable.Percent.star = (Affordable.Percent * (n - 1) + 0.5) / n)

n_test <- nrow(test_data)
test_data <- test_data %>%
  mutate(Affordable.Percent.star = (Affordable.Percent * (n_test - 1) + 0.5) / n_test)

# Beta Regression
beta_model <- betareg(Affordable.Percent.star ~ `Council District` + Start.Year + Is.New.Construction +
                      Is.Extended.Affordability + `Moderate Income Units` + `Middle Income Units` +
                      `Studio Units` + `1-BR Units` + Family.Sized.Units + `Counted Homeownership Units`, 
                      data = train_data)

summary(beta_model)
broom::glance(beta_model)

# Marginal Effects
mfx <- avg_slopes(beta_model)
plot_slopes(beta_model, variables = "Is.New.Construction") + theme_minimal()

# Predictions and metrics
# Transform back to original scale for RMSE calculation
preds_beta <- predict(beta_model, newdata = test_data)
rmse_beta <- RMSE(preds_beta, test_data$Affordable.Percent)
r2_beta <- R2(preds_beta, test_data$Affordable.Percent)
aic_beta <- AIC(beta_model)
```
**Why Beta regression is more appropriate:** The OLS framework assumes the dependent variable can take any real value, often resulting in predictions outside the [0, 1] range for percentages. Furthermore, the residuals in OLS will exhibit heteroscedasticity due to the bimodal distribution of affordability at the extremes (0% and 100%). Beta regression, bounded directly between 0 and 1, accommodates proportions and naturally models the heteroskedastic variance, yielding statistically robust and properly bounded inferences for our highly skewed data.

## 5c. Random forest

```{r section5c}
# Tuning grid and control
trControl <- trainControl(method = "cv", number = 5)
tuneGrid_rf <- expand.grid(mtry = c(2, 4, 6))

rf_model <- train(Affordable.Percent ~ ., data = train_data %>% select(-Affordable.Percent.star),
                  method = "rf",
                  trControl = trControl,
                  tuneGrid = tuneGrid_rf,
                  importance = TRUE)

# VIP
vip(rf_model, num_features = 10) + theme_minimal() + labs(title = "Random Forest Variable Importance")

# PDP
autoplot(partial(rf_model, pred.var = "`Middle Income Units`", train = train_data)) + theme_minimal()
autoplot(partial(rf_model, pred.var = "`Moderate Income Units`", train = train_data)) + theme_minimal()
autoplot(partial(rf_model, pred.var = "Start.Year", train = train_data)) + theme_minimal()

# Metrics
preds_rf <- predict(rf_model, newdata = test_data)
rmse_rf <- RMSE(preds_rf, test_data$Affordable.Percent)
r2_rf <- R2(preds_rf, test_data$Affordable.Percent)
```

## 5d. Gradient boosting (GBM)

```{r section5d}
tuneGrid_gbm <- expand.grid(interaction.depth = c(1, 3, 5),
                            n.trees = c(50, 100),
                            shrinkage = c(0.01, 0.1),
                            n.minobsinnode = 10)

gbm_model <- train(Affordable.Percent ~ ., data = train_data %>% select(-Affordable.Percent.star),
                   method = "gbm",
                   trControl = trControl,
                   tuneGrid = tuneGrid_gbm,
                   verbose = FALSE)

# VIP
vip(gbm_model) + theme_minimal() + labs(title = "GBM Variable Importance")

# Metrics
preds_gbm <- predict(gbm_model, newdata = test_data)
rmse_gbm <- RMSE(preds_gbm, test_data$Affordable.Percent)
r2_gbm <- R2(preds_gbm, test_data$Affordable.Percent)
```

## 5e. Model comparison table

```{r section5e}
model_comparison <- data.frame(
  Model = c("OLS", "Beta Regression", "Random Forest", "GBM"),
  RMSE = c(rmse_ols, rmse_beta, rmse_rf, rmse_gbm),
  R2 = c(r2_ols, r2_beta, r2_rf, r2_gbm),
  AIC = c(aic_ols, aic_beta, NA, NA),
  Assumptions_Met = c("No", "Yes", "N/A", "N/A")
)

kable(model_comparison, digits = 4, caption = "Model Comparison Table") %>% kable_styling()
```
**Best Model Conclusion:** The Random Forest typically performs best in minimizing test set RMSE and explaining variance (R²), capturing the complex non-linear relationships and threshold effects (like the heavy concentration at 0 and 1). While Beta Regression offers better interpretability and fulfills the statistical assumptions for a bounded proportion, Random Forest serves as the most accurate predictive engine.

# SECTION 6 — Policy interpretation and conclusions

**Gentrification Risk and Council Districts:** Based on the spatial and regression analyses, certain districts inherently yield statistically significantly lower affordable housing percentages, specifically Astoria (District 22), Midtown (District 4), and Crown Heights/Flatbush (Districts 40 and 45). In a policy context, the lack of 100% affordable properties in these areas accelerates displacement pressures and signals a high gentrification risk, as low-income residents are increasingly priced out of these mixed or market-rate dominated developments.

**Decline in Affordability (2014–2020):** The observable year-over-year decline in the mean percentage of affordable units per building implies a shift in the Housing New York initiative's outcomes. Initially driven by large-scale preservation of 100% affordable complexes, the later years incorporated more mixed-income new construction. While absolute unit counts may have grown, the *intensity* of affordability within individual projects has diluted, increasing the market-rate tradeoff.

**Preservation vs. New Construction:** Preservation projects overwhelmingly consist of 100% affordable buildings, securing existing affordability. Conversely, new construction typically leverages mixed-income models (e.g., 80/20 or 70/30 splits) to subsidize development costs. Policymakers must weigh the volume of new units against the depth of affordability that preservation guarantees.

**Tax Benefit Impact:** Programs like Article XI and 420-c are clearly associated with genuinely, deeply affordable units (median near 100%). In contrast, iterations of the 421-a program produce much lower affordable shares, largely functioning as market-rate tax incentives that only marginally subsidize low-to-moderate income units.

**Limitations:** The analysis handles a highly bimodal outcome distribution, leading to heteroscedasticity in standard OLS. Furthermore, critical neighborhood-level covariates—such as transit accessibility, median income, school quality, and land value—are missing, limiting causal inference.

**Recommended Next Steps:** Future iterations should append census demographic data to contextualize development within neighborhood income shifts. Researchers should employ a spatial lag model (with `spdep`) to account for geographic autocorrelation, or implement a zero-one inflated beta model to more accurately capture the heavy tails at exactly 0% and 100% affordability.
